Community-Conditioned Language Models

### 2021-05-13 paper/paper.tex
update CCLM results section

We also consider performance stratified by community; that is,
$\Ppl_{M_j}$ and $\IG_{M_j}$, 
where $M_j$ is the set of messages originating from community $c_j$ 
(\cref{fig:comm-stratified-box}).
We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{3.67} to
\num{93.58} for the best LSTM model . %TODO make long table
While the absolute performance ($\Ppl_{M_j}$) of the LSTM models is better,
we jhat the best transformer models have somewhat higher information gain
than their LSTM counterparts.

\newcommand{\modelboxplot}[3]{
  \addplot+[
    boxplot={draw position=#3, box extend=0.3}, 
    draw=#2, mark=*, mark options={fill=#2, scale=0.5}, solid, fill=#2!10,
    area legend] 
    table [y=#1] {floats/comm.csv};
}
\begin{figure}[t]
\begin{tikzpicture}
  \begin{axis}[name=ig,
    height=7cm,width=8cm,
    boxplot/draw direction=x, axis x line*=bottom, axis y line*=left,y axis line style={opacity=0},
    %xmax=1.6,
    ylabel=$l_c$, xlabel=$\IG_{M_j}$,
    legend style={font=\tiny,at={(1,1)},anchor=north east}, legend columns=2
    ]
    \addplot[thin, red, samples=2] coordinates {(1,-0.5)(1,3.5)};
    \foreach \lc in {0, ..., 3}{
      \modelboxplot{lstm-3-\lc_ig}{violet}{\lc+0.2}
      \modelboxplot{transformer-3-\lc_ig}{teal}{\lc-0.2}
    \legend{,LSTM, Transformer}
    }
  \end{axis}
  \begin{axis}[name=ppl,
    anchor=north west,at={($(ig.south west)-(0,1.1cm)$)},
    width=8cm, height=2.5cm, 
    boxplot/draw direction=x, axis x line*=bottom, hide y axis,
    xlabel=$\Ppl_{M_j}$,
    ]
    \modelboxplot{lstm-3_ppl}{violet}{0.2}
    \modelboxplot{transformer-3_ppl}{teal}{-.2}
  \end{axis}
\end{tikzpicture}
\caption{
   Average model performance by community. 
   Also see table 1 in the supplementary materials
   for a tabular version of these results with community names.
}
\label{fig:comm-stratified-box}
\end{figure}

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-info-gain}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are /r/counting, /r/friendsafari, and /r/Fireteams, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly strucutred and regular formats, which 
  are presumably conventional to these communities.}

The variation in information gain across communities is not 
related to perplexity, however (\cref{fig:ppl-info-gain}).

\begin{figure}
\begin{tikzpicture}
\begin{axis}[%
    axis x line*=bottom, axis y line*=left,
    xlabel=$\Ppl_{M_j}$, ylabel=$\IG_{M_j}$,
    width=7cm,
  ]
  \addplot+[%
    only marks,
    mark=*,
    mark options={scale=0.35, solid, color=violet!80, fill=violet},
    font=\tiny,
  ] table [x={lstm-3_ppl}, y={lstm-3-3_ig}] {floats/comm.csv};
\end{axis}
\end{tikzpicture}
\caption{
  Average community perplexity (x-axis) versus information gain (y-axis)
  for the best LSTM model ($l_c = 1$).
}
\label{fig:ppl-info-gain}
\end{figure}

@@

\section{Latent language model-based community classifiers}

@@

We can characterize the quality the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal dominates the matrix: the LMCC generally finds the right community for a message (\cref{fig:confusion}).

\begin{figure}
  \includegraphics[width=\linewidth]{floats/confusion_lstm-3-1}
\caption{
  Log-probability LMCC confusion matrices for the LSTM model with $l_c=1$. 
  Communities are sorted by indiscernability.}
\label{fig:confusion}
\end{figure}

But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (46 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTM
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.96$ and $p < 0.0001$.

However, we observe that community-wise perplexity is only weakly (if at all) 
correlated with community indiscernibility under the corresponding latent 
LMCC (see \cref{tab:ind-corrs} and, for example, \cref{fig:cclm-lmcc-ppl}).
In other words, the linguistic complexity of a given community
is a poor predictor of how distinctive its language is.
%
\begin{table}
  \small
  \centering
  \input{floats/ind_corrs.tex}
  \caption{ Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$).  }
  \label{tab:ind-corrs}
\end{table}

\begin{figure*}
  \begin{tikzpicture}
  \begin{groupplot}[
    group style={group size=2 by 1},
    legend style={anchor=north, legend columns=1, font=\tiny},
    every node near coord/.append style={font=\sffamily\tiny},%
    axis x line*=bottom, axis y line*=left,
  ]
  \nextgroupplot[xlabel=$\Ppl_{M_j}$, ylabel=$\Ind_j$]
      \addplot+[%
        only marks,
        mark=*,
        mark options={scale=0.35, solid, color=violet!80, fill=violet},
        font=\tiny,
      ] table [x={lstm-3-1_ppl}, y={lstm-3-1_indisc}] {floats/comm.csv};
  \nextgroupplot[xlabel=$\IG_{M_j}$, ylabel=$\Ind_j$]
      \addplot+[%
        only marks,
        mark=*,
        mark options={scale=0.35, solid, color=violet!80, fill=violet},
        font=\tiny,
      ] table [x={lstm-3-1_ig}, y={lstm-3-1_indisc}] {floats/comm.csv};
  \end{groupplot}
  \end{tikzpicture}
  \caption{%
    Community indiscernibility (y-axis), 
    in relationship to linguistic complexity (left) and 
    CCLM information gain (right). Shown here for the LSTM
    model with $l_c=1$. See \cref{tab:ind-corrs} for 
    Pearson's r correlations for all models.
  }
  \label{fig:cclm-lmcc-ppl}
\end{figure*}

@@

Computing \(d(L,S)\) is known as the orthogonal Procrustes problem~\citep{Gower2004}. The
solution is
\[d(L,S) = \mathsf {Tr} Σ\] where the matrix \(Σ\) is obtained by the singular value
decomposition \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). Each
singular \(\sigma_i\) (elements of the diagonal matrix \(Σ\)), give a measure of how much
correlation there is between the directions \(U_i\) and \(V_i\).

Further, we can apply a permutation to \(U\),\(V\) and \(Σ\) such that \(σ_i < σ_j\) iff \(i < j\). That
is, the largest singular value \(σ_0\) then corresponds to the principal
directions of correlation (\(U_0\), \(V_0\)), \(σ_1\) to the second
principal direction, etc.

The \(d(L,S)\) metric range from \(0\) (corresponding to perfect
independence) to \(n\) (corresponding to perfect correlation, obtained
for example if \(L=S\)), where $n=510$ is the number of communities
considered.

To get a more precise sense of what value of \(d(L,S)\) indicates significant correlation,
We have computed experimentally \(d(L,S)\) for 
\(L\) and \(S\) chosen as random 16 dimensions. We find
a mean of TODO and a standard deviation of TODO.

TODO: !!! we should take a 16 × 510 and a 300 × 510 matrix to test!!!

Thus if \(d(L,S)\) is below \(μ-3σ\), we can safely assume that there
is statistically significant correlation between \(L\) and
\(S\). However, below, we will also give a sense of \emph{how} the
correlation is manifested by analysis of the two principal components
of correlation in the linguistic embeddings, \(U_0\) and \(U_1\). To
do so we will plot the projection of embeddings on along \(U_0\) and
\(U_1\), together with the corresponding singular values.

>>>>

We also consider performance stratified by community; that is,
$\Ppl_{M_j}$ and $\IG_{M_j}$, 
where $M_j$ is the set of messages originating from community $c_j$ 
(\cref{fig:comm-stratified-box}).
We observe a lot of variation in baseline perplexity
across communities, with $\Ppl_{M_j}$ ranging from \num{3.67} to
\num{93.58} for the best LSTM model . %TODO make long table
While the performance ($\Ppl_{M_j}$) of the LSTM models is better,
we find that the best transformer CCLM has higher information gain 
than the best LSTM.

\newcommand{\modelboxplot}[3]{
  \addplot+[
    boxplot={draw position=#3, box extend=0.3}, 
    draw=#2, mark=*, mark options={fill=#2, scale=0.5}, solid, fill=#2!10,
    area legend] 
    table [y=#1] {floats/comm.csv};
}

\begin{figure}[t]
\begin{tikzpicture}
  \begin{axis}[name=ig,
    height=7cm,width=8cm,
    boxplot/draw direction=x, axis x line*=bottom, axis y line*=left,y axis line style={opacity=0},
    ylabel=$l_c$, xlabel=$\IG_{M_j}$,
    legend style={font=\tiny,at={(1,1)},anchor=north east}, legend columns=2
    ]
    \addplot[thin, red, samples=2] coordinates {(1,-0.5)(1,3.5)};
    \foreach \lc in {0, ..., 3}{
      \modelboxplot{lstm-3-\lc_ig}{violet}{\lc+0.2}
      \modelboxplot{transformer-3-\lc_ig}{teal}{\lc-0.2}
    \legend{,LSTM, Transformer}
    }
  \end{axis}
  \begin{axis}[name=ppl,
    anchor=north west,at={($(ig.south west)-(0,1cm)$)},
    width=8cm, height=2.5cm, 
    boxplot/draw direction=x, axis x line*=bottom, hide y axis,
    xlabel=$\Ppl_{M_j}$,
    ]
    \modelboxplot{lstm-3_ppl}{violet}{0.2}
    \modelboxplot{transformer-3_ppl}{teal}{-.2}
  \end{axis}
\end{tikzpicture}
\caption{
   Average model performance by community. 
   Also see table 1 in the supplementary materials
   for a tabular version of these results with community names.
}
\label{fig:comm-stratified-box}
\end{figure}

The conditioned models also perform differently across different communities---%
for some communities, $\IG_{M_j}$ is around \num{1}, or even slightly lower,
meaning that the CCLM has no advantage over the unconditioned baseline.
For other communities it is much higher, meaning that the CCLM performs better 
(\cref{fig:comm-info-gain}).\footnote{
  Some of the outlier communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English.}

This variation in information gain across communities is not 
related to perplexity, however. 
Some of the most linguistically complex communities benefit a lot
from community conditioning, while others do not benefit at all. 
Similarly, among the less linguistically complex communities 
there some have high information gain, and some have information
gain of \num{1} or less (\cref{fig:ppl-info-gain}).

%\begin{figure}
%\begin{tikzpicture}
%\begin{axis}
%\commscatter[]{lstm-3-1_ppl}{lstm-3-1_ig}
%\end{axis}
%\end{tikzpicture}
%\caption{
  %Average community perplexity (x-axis) versus information gain (y-axis)
  %for the LSTM model with $l_c = 1$ (colors based on \cref{fig:comm-pca}).
%}
%\label{fig:ppl-info-gain}
%\end{figure}

@@

\subsection{Latent language model-based community classifiers}

@@

We can characterize the quality the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test set) message $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal dominates the matrix: the LMCC generally finds the right community for a message (\cref{fig:confusion}).
% TODO: new results
%\begin{figure}
%\begin{tikzpicture}
  %\confusionsingleplot{lstm-3-1}
%\end{tikzpicture}
%\caption{LMCC confusion matrices for the LSTM model with $c=1$.}
%\label{fig:confusion}
%\end{figure}
%
But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognize messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (46 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTM
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.98$ and $p < 0.0001$ (see, for example,
\cref{fig:indisc-corr}).
%
%\begin{figure}
  %\begin{tikzpicture}
    %\begin{axis}
    %\commscatter{lstm-3-1_indisc}{transformer-3-3_indisc}
    %\end{axis}
  %\end{tikzpicture}
  %\caption{%
    %Correlation between indiscernibility for 
    %LSTM ($c = 1$) and transformer ($c = 3$) models
    %(Pearson's $r = 0.99$, $p < 0.0001$).
  %}
  %\label{fig:indisc-corr}
%\end{figure}
%
However, we observe that community-wise perplexity is only weakly (if at all) 
correlated with community indiscernibility under the corresponding latent 
LMCC (see \cref{tab:ind-corrs} and, for example, \cref{fig:cclm-lmcc-ppl}).
In other words, the linguistic complexity of a given community
is a poor predictor of how distinctive its language is.
%
\begin{table}
  \small
  \centering
  \input{floats/ind_corrs.tex}
  \caption{ Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$).  }
  \label{tab:ind-corrs}
\end{table}
%
%\begin{figure*}
  %\begin{tikzpicture}
  %\begin{groupplot}[
    %group style={group size=2 by 1},
    %legend style={anchor=north, legend columns=1, font=\tiny},
    %every node near coord/.append style={font=\sffamily\tiny}%
  %]
    %\nextgroupplot
    %\commscatter[]{lstm-3-1_ppl}{lstm-3-1_indisc}
    %\nextgroupplot
    %\commscatter[]{lstm-3-1_ig}{lstm-3-1_indisc}
    %\legend{General interest,Videogames,Technology,Sports,Female-focused,Other}
  %\end{groupplot}
  %\end{tikzpicture}
  %\caption{%
    %Average community indiscernibility (y-axis), 
    %in relationship to linguistic complexity (left) and 
    %CCLM information gain (right). Shown here for the LSTM
    %model with $l_c=1$. See \cref{tab:ind-corrs} for 
    %Pearson's r correlations for all models.
  %}
  %\label{fig:cclm-lmcc-ppl}
%\end{figure*}

@@

Computing this distance is known as the orthogonal Procrustes problem~\citep{Gower2004}. Its
solution is
\[d(L,S) = \mathsf {Tr} Σ\] where the matrix \(Σ\) is obtained by the singular value
decomposition \(U^TΣV = LS^T\). Additionally the vectors of \(U\) and \(V\) give
the directions of correlation respectively of \(L\) and \(S\). The
singular values \(\sigma_i\) (diagonal elements of \(Σ\)), say how much
correlation there is between the directions \(U_i\) and \(V_i\).

Further, we can arrange the solution so that \(σ_i < σ_j\) iff \(i < j\). That
is, the largest singular value \(σ_0\) corresponds to the principal
directions of correlation (\(U_0\), \(V_0\)), \(σ_1\) to the second
principal direction, etc.

The \(d(L,S)\) metric varies between \(0\) (perfect independence) and
\(n\) (perfect correlation, obtained for example if \(L=S\)), where
$n$ is the number of community embeddings considered.

To get a more precise sense of what value of \(d(L,S)\) indicates significant correlation,
We have computed experimentally \(d(L,S)\) for 
\(L\) and \(S\) chosen at random and 16 dimensions. We find
a mean of TODO and a standard deviation of TODO.

Thus \(d(L,S)\) below \(μ-3σ\) can be safely considered to indicate a
statistically significant correlation between \(L\) and
\(S\). However, below, we will also give a sense of \emph{how} the
correlation is manifested by analysis of the two principal components
of correlation in the linguistic embeddings, \(U_0\) and \(U_1\).

### 2021-05-13 paper/plot_macros.tex
update CCLM results section

\newcommand{\commscatter}[3][\community]{%
  \addplot[
  scatter, only marks,
  font=\tiny,
  scatter src=explicit symbolic,
  scatter/classes={
    general-interest={mark=*, blue},
    videogames={mark=*, red},
    technology={mark=*, green},
    sports={mark=*, cyan},
    female-focused={mark=*, magenta},
    other={mark=o, black}%
  },
  nodes near coords*={#1},
  nodes near coords align={vertical},
  every node near coord/.append style={font=\sffamily\tiny},
  visualization depends on={value \thisrow{community} \as \community},
  ] table [x={#2}, y={#3}, meta=category] {floats/comm.csv};
}

@@

\newcommand{\pairwisecommsim}[2]{%
  \begin{tikzpicture}
    \begin{axis}[]
      \addplot[
        scatter, only marks,
        font=\tiny,
        scatter src=explicit symbolic,
        scatter/classes={
          general-interest={mark=*, blue},
          videogames={mark=*, red},
          technology={mark=*, green},
          sports={mark=*, cyan},
          female-focused={mark=*, magenta},
          other={mark=o, black},
          different={mark=x}
        },
        ] table [x=#1, y=#2, meta=category] {floats/comm_comm.csv};
    \end{axis}
  \end{tikzpicture}
}

\newcommand{\infogainboxplot}{
  \begin{groupplot}[
    group style={
      group size=1 by 2,
      every plot/.style={
        xmin=0.65,
        xmax=1.25,
      },
    },
    height=4cm, width=8cm,
    ytick={1,2,3,4,5,6,7,8},
    yticklabels={0, 1, 2, 3, 0, 1, 2, 3},
    ylabel={$l_c$},
    ]
  \nextgroupplot
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-0_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-1_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-2_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-3_info_gain] {floats/comm.csv};
  \nextgroupplot
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-0_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-1_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-2_info_gain] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-3_info_gain] {floats/comm.csv};
  \end{groupplot}
}

>>>>

\newcommand{\commscatter}[3][\community]{%
  \addplot[
  scatter, only marks,
  font=\tiny,
  scatter src=explicit symbolic,
  nodes near coords*={#1},
  nodes near coords align={vertical},
  every node near coord/.append style={font=\sffamily\tiny},
  visualization depends on={value \thisrow{community} \as \community},
  ] table [x={#2}, y={#3}] {floats/comm.csv};
}

@@

\newcommand{\pairwisecommsim}[2]{%
  \begin{tikzpicture}
    \begin{axis}[]
      \addplot[
        scatter, only marks,
        font=\tiny,
        scatter src=explicit symbolic,
        ] table [x=#1, y=#2] {floats/comm_comm.csv};
    \end{axis}
  \end{tikzpicture}
}

\newcommand{\infogainboxplot}{
  \begin{groupplot}[
    group style={
      group size=1 by 2,
      every plot/.style={
        xmin=0.65,
        xmax=1.25,
      },
    },
    height=4cm, width=8cm,
    ytick={1,2,3,4,5,6,7,8},
    yticklabels={0, 1, 2, 3, 0, 1, 2, 3},
    ylabel={$l_c$},
    ]
  \nextgroupplot
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-0_ig] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-1_ig] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-2_ig] {floats/comm.csv};
  \addplot+[boxplot,draw=blue,mark=*,solid]
  table [y=lstm-3-3_ig] {floats/comm.csv};
  \nextgroupplot
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-0_ig] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-1_ig] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-2_ig] {floats/comm.csv};
  \addplot+[boxplot,draw=red,mark=*,solid]
  table [y=transformer-3-3_ig] {floats/comm.csv};
  \end{groupplot}
}

% https://tex.stackexchange.com/questions/183778/how-can-several-boxplots-be-combined-into-a-single-figure-as-groups/183856#183856
